Forecasting Gamma-Ray Bursts using Gravitational Waves

### 2018-06-11 Paper/Manuscript.tex
Change of reference
Added 3 new references and text accordingly

Our computation for $T_\text{AW}$ is based on the Newtonian evolution of a binary system composed of two point masses in a quasi-circular Keplerian orbit. 
The long-term dynamics is governed by the radiation reaction of 
the GW emission from the system due to the time variance of the quadrupole moment of the binary.
Though the Newtonian (leading-order) treatment gives the 
inspiral time with better than $ 95\%$ accuracy,
we nonetheless supplement it with post-Newtonian corrections up to 3.5 post-Newtonian order (3.5PN) \cite{Blanchet_LRR}.
However, even that fails to provide an accurate depiction of the binary
evolution in the strong-field regime, where speeds $\gtrsim 0.1c$. This is a very complex and complicated problem requiring fully
numerical treatment in terms of general relativistic, magnetohydrodynamic codes 
with neutrino transport running on large-scale computing clusters and taking up to millions of CPU hours 
(see Ref.~\cite{Faber:2012rw} for a review, and Refs.~\cite{Baiotti:2016qnr,Kyutoku:2017voj, Zappa:2017xba, Dietrich:2018upm} for the latest developments and references therein).
Despite its simple formulation, our quadrupole radiation driven Newtonian
evolution is more than sufficient to provide reliable estimations of advance warning times.
We support this claim in Sec.~\ref{sec:idealizations} with an extensive list of computations showing how much each of our approximations affects the inspiral time.

@@

the fastest observed NS spin $\lesssim 0.04$ \cite{FastestNSspin}.

@@

\section{GW170817 as a test of our model for the Advanced LIGO-Virgo network}\label{sec:GW170817}
On 17 August 2017, at 12:41:04 UTC, GWs from the inspiral and merger of a binary neutron star system %were detected by the Advanced LIGO-Virgo network.
%The event, first picked up by LIGO Hanford (H1), 
swept the ALV network's frequency band from $\approx 30\,$Hz to $\approx 2000\,$Hz, accumulating a total SNR of
%18.8, 26.4 and 2.0 at H1, LIGO Livingstone (L1) and Virgo (V1), respectively, yielding 
$\rho_\text{tot} = 32.4$ \cite{GW170817}.
The inferred duration of the event in the network bandwidth was $\approx 57\,$ seconds during which $\approx 3000$ cycles of GWs were emitted \cite{GW170817_Facts}.
Assuming low-spin priors  ($\chi \le 0.05$, see point 7 of Sec.~\ref{sec:idealizations}), the following values were inferred for the total mass, the chirp mass,
and the luminosity distance of the source with 90\% confidence:
%
\be
M = 2.74^{+0.04}_{-0.01} M_\odot,\quad M_c = 1.188^{+0.004}_{-0.002} M_\odot,\quad D= 40^{+8}_{-14}\,\text{Mpc}. \label{eq:GW170817_params}
\ee
%
Note that the error bar for the chirp mass is very small
due to the fact that it can be extracted to high precision over thousands of GW cycles via Eq.~(\ref{eq:fdot}).

@@

We now put our model three-IFO network to the test
by computing the SNRs accumulated by each individual IFO and by the entire network. To do so, we take the best-inferred values for $M,M_c,D$ from GW170817 given in Eq.~(\ref{eq:GW170817_params}) above and compute the SNRs $\rho_i, i=h,m,l$ via a slightly modified version of Eq.~(\ref{eq:SNR_i})
%%
%
\begin{align}
\bar{\rho}_i\equiv\rho_i(f_0,f_f) &= 2\tilde{A}_i\, h_0\left[ \int_{f_0}^{f_f} df\, \f{f^{-7/3}}{S_\text{n}(f)}\right]^{1/2} \label{eq:SNR_i_with_fi}.
\end{align}
%
where $f_0 =30\,\text{Hz}, f_f= f_\text{ISCO}$ for GW170817 and $\tilde{A}_i$ given in Table~\ref{table:network_params}.
For this computation, we have neglected the cosmological corrections as 40\,Mpc is in our ``local'' neighbourhood.

@@

There is additional motivation to choose low BH mass as 
 $\tau_\text{insp} \propto M_c^{-5/3} = (1+Q)^{1/3}/Q$ for $Q\equiv M_\text{BH}/M_\text{NS} > 1$, which scales as $Q^{-2/3}$ as $Q\to\infty$ , i.e., heavier binaries inspiral faster
 so it is desirable to minimize $Q$.
 Setting $M_\text{NS}=1.4 M_\odot$ as before gives $Q=5/1.4  \simeq 3.571$. $c_\text{NS}$ is determined by the radius of the NS constrained to be in the $9.9 - 11.2\,$km range \cite{Ozel:2016oaf}
 which results in $0.13\lesssim c_\text{NS} \lesssim 0.15$ 
 for our chosen NS mass.
%As we wish to avoid discussing the details of the strong-field evolution of the binary, we borrow certain results from the literature to
%justify our choice for the chosen parameters above. 
If we pick the more pessimistic $c_\text{NS} = 0.15$ bound then we must have $Q\lesssim 5$ and $a\gtrsim 0.5$ (prograde) for TD to occur \cite{Ferrari:2008nr}
and $Q< 3$ in the case of $a=0$ \cite{Kyutoku:2011vz}.

>>>>

Our computation for $T_\text{AW}$ is based on the Newtonian evolution of a binary system composed of two point masses in a quasi-circular Keplerian orbit. 
The long-term dynamics is governed by the radiation reaction of 
the GW emission from the system due to the time variance of the quadrupole moment of the binary.
Though the Newtonian (leading-order) treatment gives the 
inspiral time with better than $ 95\%$ accuracy,
we nonetheless supplement it with post-Newtonian corrections up to 3.5 post-Newtonian order (3.5PN) \cite{Blanchet_LRR}.
However, even that fails to provide an accurate depiction of the binary
evolution in the strong-field regime, where speeds $\gtrsim 0.1c$. This is a very complex and complicated problem requiring fully
numerical treatment in terms of general relativistic, magnetohydrodynamic codes 
with neutrino transport running on large-scale computing clusters and taking up to millions of CPU hours 
(see Ref.~\cite{Faber:2012rw} for a review, and Refs.~\cite{Baiotti:2016qnr,Kyutoku:2017voj, Zappa:2017xba, Dietrich:2018upm, Dietrich:2018phi} for the latest developments and references therein).
Despite its simple formulation, our quadrupole radiation driven Newtonian
evolution is more than sufficient to provide reliable estimations of advance warning times.
We support this claim in Sec.~\ref{sec:idealizations} with an extensive list of computations showing how much each of our approximations affects the inspiral time.

@@

the fastest observed NS spin in a BNS system is $\lesssim 0.04$ \cite{0264-9381-26-7-073001}.

@@

\section{GW170817 as a test of our model for the Advanced LIGO-Virgo network}\label{sec:GW170817}
On 17 August 2017, at 12:41:04 UTC, GWs from the inspiral and merger of a binary neutron star system %were detected by the Advanced LIGO-Virgo network.
%The event, first picked up by LIGO Hanford (H1), 
swept the ALV network's frequency band from $\approx 30\,$Hz to $\approx 2000\,$Hz, accumulating a total SNR of
%18.8, 26.4 and 2.0 at H1, LIGO Livingstone (L1) and Virgo (V1), respectively, yielding 
$\rho_\text{tot} = 32.4$ \cite{GW170817}.
The inferred duration of the event in the network bandwidth was $\approx 57\,$ seconds during which $\approx 3000$ cycles of GWs were emitted \cite{GW170817_Facts}.
Assuming low-spin priors  ($\chi \le 0.05$, see point 7 of Sec.~\ref{sec:idealizations}), the following values were inferred for the total mass, the chirp mass,
and the luminosity distance of the source with 90\% confidence:
%
\be
M = 2.74^{+0.04}_{-0.01} M_\odot,\quad M_c = 1.188^{+0.004}_{-0.002} M_\odot,\quad D= 40^{+8}_{-14}\,\text{Mpc}. \label{eq:GW170817_params}
\ee
%
Note that the error bar for the chirp mass is very small
due to the fact that it can be extracted to high precision over thousands of GW cycles via Eq.~(\ref{eq:fdot}). In fact, detailed follow-up analyses (\cite{GW170817_2018}) have recently restricted the above-values to
%
$ \left\{2.73^{+0.04}_{-0.01} M_\odot, 1.186^{+0.001}_{-0.001} ,  40.7^{+2.36}_{-2.36}\text{Mpc}\right\}$, respectively,
%
where the errors in $D$ are at $1\sigma$ level.

@@

We now put our model three-IFO network to the test
by computing the SNRs accumulated by each individual IFO and by the entire network. To do so, we take the originally-inferred values for $M,M_c,D$ from GW170817 given in Eq.~(\ref{eq:GW170817_params}) above and compute the SNRs $\rho_i, i=h,m,l$ via a slightly modified version of Eq.~(\ref{eq:SNR_i})
%%
%
\begin{align}
\bar{\rho}_i\equiv\rho_i(f_0,f_f) &= 2\tilde{A}_i\, h_0\left[ \int_{f_0}^{f_f} df\, \f{f^{-7/3}}{S_\text{n}(f)}\right]^{1/2} \label{eq:SNR_i_with_fi}.
\end{align}
%
where $f_0 =30\,\text{Hz}, f_f= f_\text{ISCO}$ for GW170817 and $\tilde{A}_i$ given in Table~\ref{table:network_params}.
For this computation, we have neglected the cosmological corrections as 40\,Mpc is in our ``local'' neighbourhood.

@@

There is additional motivation to choose low BH mass as 
 $\tau_\text{insp} \propto M_c^{-5/3} = (1+Q)^{1/3}/Q$ for $Q\equiv M_\text{BH}/M_\text{NS} > 1$, which scales as $Q^{-2/3}$ as $Q\to\infty$ , i.e., heavier binaries inspiral faster
 so it is desirable to minimize $Q$.
 Setting $M_\text{NS}=1.4 M_\odot$ as before gives $Q=5/1.4  \simeq 3.571$. $c_\text{NS}$ is determined by the radius of the NS constrained to be in the $9.9 - 11.2\,$km range \footnote{Recent work by the ALV collaboration \cite{GW170817_radii}) give bounds of $R_\text{NS}=9.1 -11.8\,$km, more or less consistent with astrophysical results.}\cite{Ozel:2016oaf}
 which results in $0.13\lesssim c_\text{NS} \lesssim 0.15$ 
 for our chosen NS mass. 
%As we wish to avoid discussing the details of the strong-field evolution of the binary, we borrow certain results from the literature to
%justify our choice for the chosen parameters above. 
If we pick the more pessimistic $c_\text{NS} = 0.15$ bound then we must have $Q\lesssim 5$ and $a\gtrsim 0.5$ (prograde) for TD to occur \cite{Ferrari:2008nr}
and $Q< 3$ in the case of $a=0$ \cite{Kyutoku:2011vz}.
